\documentclass{article}

\usepackage{xspace}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{listings}

\newcommand{\thetool}{{\it FairTest}\xspace}
\newcommand{\heading}[1]{\noindent{\bf{#1}}}
\newcommand{\xxx}[1]{\textbf{{XXX} #1}}

\title{\thetool: A Unit Testing Framework for Uncovering Privacy Bugs in
  Internet-of-Things Applications}
\author{V. Atlidakis, D. X. He, E. D. Park}
\date{\today}

\begin{document}
\maketitle

\heading{Motivation.}
The Internet of Things (IoT) offers enormous capabilities for data exchange
enhancing societal progress, but it comes with a strong need for transparency
and accountability with respect to data usage. All the more, highly sensitive
data are transmitted from devices on our bodies, in our homes, and in the
streets. Unfortunately, the privacy implications of collecting and using that
information are poorly understood and rarely studied. Consequently, there is a
clear danger of unintentional algorithmic discriminations. An indicative recent
example is the Staples price discrimination case~\cite{Staples}.
In this case an arguably reasonable algorithmic decision to optimize online
prices based on user proximity to competitor brick-and-mortar locations led to
consistently higher prices for low-income populations, since they tend to
generally live farther from these stores. This kind of indirect--and likely
unintended--effects are challenging to identify, and with the use of sensitive
information collected by IoT applications, the risk for these dangers is only
increasing. Thus, we believe that new programming tools are needed to increase
developers’ visibility into the data they collect and the implications of its
integration into applications.

\heading{Goals.}
Our purpose is to build \thetool: A discrimination testing suite uncovering
differential treatment of various populations based on protected
attributes, such as religion, sexual orientation, or income. Our idea is to
measure correlation between outputs of a specific program (such as a price
recommendation engine) and the values of the protected attributes. Any strong
correlation between protected attributes and outputs is threated as a privacy
bug and is reported to the
developer as potential discrimination. To provide a better perspective to the
programmer, we will try to develop meaningful heuristics that rank observed
correlations based on their likelihood to be unintended or unfair
discriminations versus natural, preference-based effects. We will design
{\it FairTest} to be easy to use by regular application programmers,
will integrate
it into popular testing frameworks, and write a set of sample applications
indicating meaningful use cases.

\heading{Related Work.}
Fairness in the use of data by IoT applications is drawing increasing
attention in our days, and yet there is no established methodology for
measuring fairness of online algorithms producing recommendations, nor is
there an established methodology for shading some light and helping
uncover potential online discrimination cases. The authors of~\cite{Fairness}
study fairness of classification and their purpose is to prevent
discriminating individuals based on certain characteristics, such as
membership in a group, and at the same time maintain the effectiveness of
the classifier. The limitation of their approach is that it requires
programmes to be able to define similarity metrics, and it is unclear
whether programmers are able to define such metrics or not. Also,
in~\cite{DisparateImpact} unequal treatment of different populations
by a classification methodology is studied for its legal concept of
disparate impact. This work formalizes what it means for a dataset to
have the potential for disparate impact if used by a classifier, and
suggests methods for detecting and removing potential disparate impact.
However, it is not showing the effectiveness of the proposed procedures
in real applications, and also does not extend to general discrimination
hazards.

\heading{Expected Conclusions.}
By using \thetool a developer will be able to test an application for
discrimination cases by uncovering correlations between protected attributes
and outputs. We expect to be able to identify clear cut discrimination cases,
as well as previously unknown obscure discrimination cases. \thetool should
work independently from the backend algorithms used for recomendations,
and it should feature a
simple API allowing developers easily test their code for privacy bugs.

\heading{Research Plan.}
The first milestone is to build a prototype of our system no later than
the project status presentation date, on 3/10. Our prototype will include a
demo dataset for an application imitating a known discrimination case,
such as the Staple~\cite{Staples} case. We expect to be able to identify
strong correlations between protected attributes and outputs in clear cut
cases such as in~\cite{Staples}. The second milestone is to augment our
prototype with real datasets (hopefully obtained from online web stores)
and a set of simple applications featuring a real price recommendation
engine. Our hope is that we will be able to uncover obscure correlations
of protected attributes and outputs that are impossible to foresee by the
programmer. The third milestone is to finalize the implementation of \thetool
and integrate it into a popular testing framework. The fourth milestone,
is to present our experiences and lessons learned as well as interesting
(and previously unknown) unintentional cases of discrimination
uncovered by \thetool.

{
  \scriptsize
  \setlength\itemsep{0pt}
  \footnotesize
  \bibliographystyle{abbrv}
  \bibliography{whitepaper}
}
\end{document}
